Finding Subgraphs With Side Constraints\thanks{This research was supported by the Engineering and Physical Sciences Research Council [grant number EP/P026842/1]}

### 2021-01-27 paper/paper.tex
More or less done

\author{
    \"Ozg\"ur Akg\"un\inst{1} \and Jessica Enright\inst{2} \and Christopher
    Jefferson\inst{1} \and Ciaran McCreesh\inst{2} \and Patrick
    Prosser\inst{3} \and Steffen Zschaler\inst{4} \\
}
\authorrunning{\"O. Akgun et al.}
\institute{
    University of St Andrews, Scotland \and
    University of Glasgow, Scotland \\ \email{ciaran.mccreesh@glasgow.ac.uk} \and
    Algorithmicists Anonymous, Scotland \and
    King's College London, England
}

@@

\begin{abstract}
    The subgraph isomorphism problem is to find a small ``pattern'' graph inside a larger ``target''
    graph. There are excellent dedicated solvers for this problem, but they require substantial
    programming effort to handle complex side constraints; however, general purpose constraint
    solvers struggle on more difficult graph instances. We show how to combine the state of the art
    Glasgow Subgraph Solver with the Minion constraint programming solver to get a ``subgraphs modulo
    theories'' solver that is both performant and flexible. We also show how such an approach can be
    driven by the Essence high level modelling language, giving ease of modelling to non-expert
    users. We give practical examples involving
    temporal graphs, typed graphs from software engineering, and costed subgraph isomorphism
    problems.
\end{abstract}

@@

Finding or counting small ``pattern'' graphs inside larger ``target'' graphs are widely applicable
hard problems, with applications including ??. This has led to the development of numerous dedicated
algorithms, with the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020} being the current
state of the art \cite{DBLP:conf/gbrpr/Solnon19}. However, practitioners are often interested in
versions of the problem with additional restrictions, or side constraints. Some of these, such as
exact vertex labelling schemes, are trivial to include in a dedicated solver, but others currently
require either extensive programming or inefficient post-processing. This paper explores a different
approach: by allowing the Glasgow Subgraph Solver to use the Minion constraint programming (CP) solver
\cite{DBLP:conf/ecai/GentJM06} for side constraints, we achieve both the performance only a
dedicated solver can offer, with the flexibility of a full CP toolkit. This
hybrid modelling system can be driven by the Essence high level modelling language
\cite{DBLP:journals/constraints/FrischHJHM08} and the Conjure toolchain, making it accessible to
non-specialists.

@@

As a simple example, the following inputs show the problem instance represented in
\cref{figure:littlegraphs}. We have three different vertex labels (circle, square, and diamond), and
only a single edge type (which is directed; the numerical labels on edges are not used in this
section):
\begin{lstlisting}
letting l be 3
letting e be 1
\end{lstlisting}
We may now describe the pattern:
\begin{lstlisting}
letting p be 3
letting pat be function ((1, 2) --> 1, (2, 3) --> 1)
letting plab be function (1 --> 1, 2 --> 1, 3 --> 2)
\end{lstlisting}
and the target:
\begin{lstlisting}
letting t be 5
letting tgt be function ((1, 3) --> 1, (3, 1) --> 1,
  (2, 3) --> 1, (3, 4) --> 1, (3, 5) --> 1)
letting tlab be function (1 --> 1, 2 --> 1, 3 --> 1,
  4 --> 2, 5 --> 3)
\end{lstlisting}
Using the Conjure tool to compile Essence to a constraint programming model which is then solved by
Minion, we find there are exactly two solutions to the problem, as we would expect:
\begin{lstlisting}
(1 --> 1, 2 --> 3, 3 --> 4)
(1 --> 2, 2 --> 3, 3 --> 4)
\end{lstlisting}
But what if our application requires induced isomorphisms? Then we can easily add the constraint
\begin{lstlisting}
such that forAll (a, b) : (P, P) .
    (f(a), f(b)) in defined(tgt) -> (a, b) in defined(pat)
\end{lstlisting}
And Conjure will now find us a single solution,
\begin{lstlisting}
(1 --> 2, 2 --> 3, 3 --> 4)
\end{lstlisting}
As we will see in \cref{section:problems}, supporting other problem variants and constraints is
similarly straightforward, even if auxiliary variables are required. %For example, if instead we
%want to allow relabelling on vertex labels (which
%is typically not supported by dedicated solvers), we could do the following:
%\begin{lstlisting}
%find r : function (total, injective) L --> L
%such that forAll a : P .  r(plab(a)) = tlab(f(a))
%\end{lstlisting}
%and we would find two additional solutions,
%\begin{lstlisting}
%(1 --> 1, 2 --> 3, 3 --> 5)
%(1 --> 2, 2 --> 3, 3 --> 5)
%\end{lstlisting}
%and if we removed the injective keyword for the relabelling, we would find a fifth mapping
%\begin{lstlisting}
%(1 --> 2, 2 --> 3, 3 --> 1)
%\end{lstlisting}
%We return to relabelling in \cref{section:typegraphs}.

@@

@@

Unfortunately, whilst elegant and flexible, the performance of this approach is rather terrible on
basic subgraph isomorphism instances.  The computational experiments in this paper are performed on
a cluster of machines with dual Intel Xeon E5-2697A v4 processors and 512GBytes RAM running Ubuntu
18.04.  Initially, we will be using the 14,621 unlabelled, undirected instances from Solnon's
benchmark suite\footnote{https://perso.liris.cnrs.fr/christine.solnon/SIP.html}. This benchmark
suite was originally designed for algorithm portfolios work \cite{DBLP:conf/lion/KotthoffMS16}, and
brings together several collections of application and randomly-generated instances with varying
difficulties and solution counts (including many unsatisfiable instances). Some of the instances
have up to 900 vertices and 14,420 edges in patterns and up to 6,671 vertices and 209,000 edges in
targets. These lead to rather large models, by constraint programming standards: the largest
generated table constraint has nearly half a million entries.

In \cref{figure:solvers} we plot the cumulative number of instances solved over time for the
non-induced decision problem, comparing the high level approach to the Glasgow Subgraph Solver
\cite{DBLP:conf/gg/McCreeshP020} and PathLAD \cite{DBLP:conf/lion/KotthoffMS16} (the two strongest
CP approaches), and to VF2 \cite{DBLP:journals/pami/CordellaFSV04} and RI
\cite{DBLP:journals/bmcbi/BonniciGPSF13} (simpler algorithms which perform well on easy instances).
The high level approach has very slow startup times (which is to be expected as it involves
launching a Java virtual machine and reading in a very large table constraint), but much more
worryingly, only catches up with the worst other solver in number of instances solved as the timeout
approaches. Worse, as the scatter plot in \cref{figure:solvers} shows, there are almost no instances
where the high level approach does better than the Glasgow Subgraph Solver. (For induced problems,
the results are even less favourable.)

These first results motivate the remainder of this paper. We want to retain the convenience of the
high level modelling approach, and to be able to add arbitrary side constraints to suit different
applications, but we do not want to have to abandon the performance that dedicated solvers can get
on hard instances. In \cref{section:hybrid} we evaluate several ways of using a CP solver in
conjunction with the Glasgow Subgraph Solver, with a focus on low level implementation details. In
\cref{section:problems} we then return to high level modelling, and look at the convenience it
provides for ??temporal problems, ??relabelling and retyping, and ??optimisation problems.

\section{Hybrid Solving}\label{section:hybrid}

The Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020} employs a CP approach to solve
subgraph-finding problems, but using special data structures and algorithms---for example, rather
than representing the adjacency constraint using a table, it uses bitset adjacency matrices
\cite{DBLP:conf/cp/McCreeshP15}. The solver also exploits various graph invariants involving degrees
\cite{DBLP:journals/constraints/ZampelliDS10} and paths \cite{DBLP:conf/cp/AudemardLMGP14} to
further reduce the search space, and employs special search order heuristics
\cite{DBLP:conf/cpaior/ArchibaldDHMP019}. From this paper's perspective, the most important design
aspect is that internally, the solver has a CP style variable for each vertex in the pattern graph,
whose domains range over the vertices of the target graph. The solver performs a backtracking search
with restarts and nogood recording, attempting to assign each variable a value from its domain,
whilst respecting adjacency and injectivity constraints. At each recursive call of search, the
solver performs \emph{propagation} to eliminate infeasible values from domains. If any domain
becomes empty, the solver backtracks; otherwise, it selects a variable, and tries assigning it each
value from its domain in turn.

For now we will assume that there is some way of setting up the subgraph solver and a CP solver such
that they both have this same set of variables and values, and that there is some way of knowing how
to form a correspondence between their internal representations. We will allow the CP solver to have
additional variables that the subgraph solver does not know about, and we do not specifically
require the CP solver to be aware of all of the graph constraints. We could therefore use a CP
solver as a \emph{solution checker}. Whenever the subgraph solver finds a solution, it will pass it
to the CP solver, which will treat the solution as a set of equality constraints. The CP solver will
then attempt to find a satisfying assignment. If the CP solver does not have any additional
variables, this is equivalent to simply checking that the remaining constraints hold, but in general
this will require search.
For a decision problem, the CP solver then communicates back to the subgraph
solver either ``yes, this is a valid solution'', or ``no, reject this solution and keep going''. If
we are solving a counting or enumeration problem, the CP solver must find \emph{all} solutions and
communicate this back to the subgraph solver.

For more power, but possibly also greater cost, we could additionally ask a CP solver at every stage
of search to test whether the state the subgraph solver is in is obviously infeasible.  Whenever the
subgraph solver has finished performing propagation, it can communicate the \emph{trail} (that is,
its current sequence of guessed assignments) to the CP solver, which again treats these as
additional equality constraints. The CP solver then performs its own propagation (but not search),
and communicates back either a ``yes, keep going'' or a ``no, backtrack immediately''.  Finally,
after this testing, we could also ask the CP solver to communicate any deletions it infers back to
the subgraph solver. In other words, the subgraph solver would use the CP solver as an additional
propagator.

@@

\subsection{Implementation}

Problems are specified in Essence and then are converted to concrete models that can be solved by
a CP solver (in our experiments Minion) using Conjure and Savile Row.
Running Conjure with a special flag (--graph-solver) instructs it to generate an extra file which
describes how the variables representing the graph is represented in the SavileRow input (known
as Essence'). This is used
by the graph solver so it can map between its internal state and the variables in Essence'.
SavileRow converts the graph model to Minion input, and this conversion contains information which
allows Minion to map its internal state back to the Essence' given to SavileRow. The graph solver
and Minion then communicate mappings between the nodes and edges in the graph using the
identifiers in the Essence' representation of the problem.

@@

We now present the results of some computational experiments. The experiments in this section are
designed to be computationally hard, and to emphasise the difference between the approaches, rather
than to be realistic. We will continue to work with Solnon's non-induced subgraph isomorphism
benchmark instances, but will consider three variations. Firstly, we will consider the problem with
no side constraints. This is, in some sense, the worst case scenario, where we must pay the full
price of hybrid solving, but cannot get any benefit from it. Secondly, let us say that the number of
odd-indexed target vertices used must be greater than the number of even target vertices used:
\newcommand{\modsymbol}{\%}
\begin{lstlisting}
such that (sum i : P . f(i) (*\modsymbol{}*) 2) >
    (sum i : P . (f(i) + 1) (*\modsymbol{}*) 2)
\end{lstlisting}
Thirdly, let us instead say that fewer than three odd target vertices may be used:
\begin{lstlisting}
such that (sum i : P . f(i) (*\modsymbol{}*) 2) < 3
\end{lstlisting}
And fourthly, let us say that even pattern vertices must be mapped to even target vertices, and odd
pattern vertices to odd target vertices:
\begin{lstlisting}
such that forAll a : P . (a (*\modsymbol{}*) 2) = (f(a) (*\modsymbol{}*) 2)
\end{lstlisting}

We chose these last three problems because the first is likely not to be able to
perform inference until deep in a search tree (when most pattern vertices are mapped to specific
target vertices), the second is likely to be able to perform inference early in the search tree
(after a few assignments have been made, but not at the root node), and the third should propagate
only at the root node.

The results of these experiments are presented in \cref{figure:cumulative}. Let us first look at the
top left plot, where we do not actually have any side constraints. When calling the CP solver as a
solution checker, we ultimately achieve the same performance as the Glasgow Subgraph
Solver\footnote{Actually, because high level modelling can use different names for variables and
values, we get slight differences due to changes to tiebreaking in search order heuristics.},
although we can pay a substantial startup overhead. This should not be surprising: with no side
constraints, the subgraph solver runs as normal, and will perform just one call to the CP solver on
satisfiable instances. The propagating approach is over an order of magnitude slower: calling the CP
solver at every search node is clearly very expensive. (We also tried calling the CP solver to test
feasibility, without communicating deletions; this made no noticable difference to performance, and
so is not pictured.) Finally, all approaches substantially outperform using a CP solver on its own
without help from the subgraph solver.

@@

As a point of comparison, we also implemented these three sets of side constraints natively inside
the Glasgow solver. For the ``more odd than even'' case, we did this through solution checking, for
``less than three odd'' we implemented checking during search, and for ``odd to odd, even to even''
we implemented initial domain filtering. Our implementation choices here are intended to reflect what a
reasonable programmer would do if the high level approach were not available (and we intentionally
selected side constraints that would not be too difficult to implement). In two of the three cases,
the hand-crafted code does somewhat outperform the hybrid solving, but in the ``more odd than even''
case, hybrid solving actually beats the hand-crafted dedicated solver implementation when using the
rollback approach, which we now describe.

@@

From what we have seen so far, it is obviously important to call the CP solver \emph{some} of the
time during search, but too expensive to call it \emph{all} of the time. We will therefore introduce
a new approach, which we call \emph{rollback}. This approach is inspired by backjumping ??cite, as
well as by the conflict analysis methods used in SAT and SMT solvers ??cite and in lazy clause
generating CP solvers ??cite. The idea is as follows.  Firstly, we call the CP solver with full
propagation at the root node, in case we are dealing with a particularly rich labelling scheme.
Secondly, we use the CP solver for solution checking, since this is required for correctness. Now,
suppose the CP solver rejects a candidate solution: this will cause the subgraph solver to
backtrack. At this point, we call the CP solver again, with full propagation. Either the CP solver
indicates feasibility, in which case we proceed with search (potentially with a reduced set of
domains), or the CP solver indicates failure, in which case we backtrack again, and do another
attempt at full propagation, and so on until feasiblity is reached.

@@

We now look at three classes of real-world subgraph-finding problems that, until now, have been
solved using dedicated approaches. We show how easy it is to model these problems in Essence. ?? do
some actual simple experiments on them?

@@

  In graph theory~\cite{Ehrig+06}, a graph is given by a set of vertices and a set of edges and two functions giving the source and target node for an edge, respectively.
  Morphisms between graphs map vertices to vertices and edges to edges, preserving the structure of the graphs.
  Often, we are interested in typed graphs.
  Here, we assume a dedicated graph TG---the type graph---to be given.
  Then we can define typed graphs as graphs for which a morphism into TG has been defined.
  In effect, we are labelling each vertex and edge with its corresponding type.

  For typed graphs, morphisms between the graphs are typically defined between graphs typed over the same type graph, but there are situations where we are interested in mapping between graphs typed over different type graphs.
  We describe an example scenario in~\cite{Duran+12,Duran+17}, where we compose graph transformation systems by defining morphisms between the rules constituting the respective transformation systems.
  In this case, the source and target transformation systems will normally have different type graphs and we will first establish a morphism between the two type graphs.
  Mappings between the various graphs making up the rules then need to preserve structure and typing subject to the morphism between type graphs.
  We have developed tool support for this approach to specification composition.\footnote{\url{https://github.com/gts-morpher/gts_morpher}}
  A key objective is to minimise the amount of specification that needs to be written.
  For example, the tool allows morphisms between graph transformation systems to be only partially specified and then automatically completes the full morphism, if it can do so unambiguously.
  This requires solving a sub-isomorphism problem.

%A \emph{typed graph} is a kind of labelled graph where the labels also carry a graph structure
%??cite. In general, mappings between typed graphs must preserve the label structure, but here we
%will be looking at a problem variation involving relabelling which occurs in ??application. 
We may describe typed graph subisomorphism problems in Essence as follows. As before, we are given a
pattern graph and a target graph, both of which carry labels; we will draw the vertex labels from
different sets, to emphasise
the relabelling.
\begin{lstlisting}
given pl, tl, e : int
letting PL be domain int(1..pl)
letting TL be domain int(1..tl)
letting E be domain int(1..e)
\end{lstlisting}
We are also given labelled graphs,
\begin{lstlisting}
given p, t : int
letting P be domain int(1..p)
letting T be domain int(1..t)

>>>>

\author{
    \"Ozg\"ur Akg\"un\inst{1}\orcidID{0000-0003-4460-6912} \and Jessica
    Enright\inst{2}\orcidID{0000-0002-0266-3292} \and Christopher
    Jefferson\inst{1}\orcidID{0000-0003-2979-5989} \and Ciaran
    McCreesh\inst{2}\orcidID{0000-0002-6106-4871} \and Patrick
    Prosser\inst{3}\orcidID{0000-0003-4460-6912} \and Steffen
    Zschaler\inst{4}\orcidID{0000-0001-9062-6637} \\
}
\authorrunning{\"O. Akgun et al.}
\institute{
    University of St Andrews, Scotland \and
    University of Glasgow, Scotland \\ \email{ciaran.mccreesh@glasgow.ac.uk} \and
    Algorithmicists Anonymous, Scotland \and
    King's College London, England
}

@@

\begin{abstract}
    The subgraph isomorphism problem is to find a small ``pattern'' graph inside a larger ``target''
    graph. There are excellent dedicated solvers for this problem, but they require substantial
    programming effort to handle the complex side constraints that often occur in practical
    applications of the problem; however, general purpose constraint solvers struggle on more
    difficult graph instances. We show how to combine the state of the art Glasgow Subgraph Solver
    with the Minion constraint programming solver to get a ``subgraphs modulo theories'' solver that
    is both performant and flexible. We also show how such an approach can be driven by the Essence
    high level modelling language, giving ease of modelling and prototyping to non-expert users. We
    give practical examples involving temporal graphs, typed graphs from software engineering, and
    costed subgraph isomorphism problems.
\end{abstract}

@@

Finding small ``pattern'' graphs inside larger ``target'' graphs is a widely applicable
hard problem, with applications including compilers \cite{DBLP:conf/cp/BlindellLCS15},
bioinformatics \cite{DBLP:journals/bmcbi/BonniciGPSF13,Elhesha2021}, chemistry \cite{o:Regin95},
malware detection \cite{DBLP:conf/dimva/BruschiMM06}, pattern recognition
\cite{DBLP:journals/ijprai/FoggiaPV14}, and the design of mechanical locks
\cite{DBLP:journals/siamrev/VomelLBF17}.  This has led to the development of numerous dedicated
algorithms, with the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020} being the current
state of the art \cite{DBLP:conf/gbrpr/Solnon19}. However, practitioners are often interested in
versions of the problem with additional restrictions, or side constraints. Some of these, such as
exact vertex labelling schemes, are trivial to include in a dedicated solver, but others currently
require either extensive programming or inefficient post-processing. This paper explores a different
approach: by allowing the Glasgow Subgraph Solver to use the Minion constraint programming (CP)
solver \cite{DBLP:conf/ecai/GentJM06} for side constraints, we achieve both the performance only a
dedicated solver can offer, with the flexibility of a full CP toolkit. This hybrid modelling system
can be driven by the Essence high level modelling language
\cite{DBLP:journals/constraints/FrischHJHM08} and the Conjure toolchain, making it accessible to
non-specialists.

@@

As a simple example, the following inputs show the problem instance represented in
\cref{figure:littlegraphs}. We have three different vertex labels (circle, square, and diamond), and
only a single edge type (which is directed; the numerical labels on edges are not used in this
section):
\begin{lstlisting}
letting l be 3
letting e be 1
\end{lstlisting}
We may now describe the pattern:
\begin{lstlisting}
letting p be 3
letting pat be function ((1, 2) --> 1, (2, 3) --> 1)
letting plab be function (1 --> 1, 2 --> 1, 3 --> 2)
\end{lstlisting}
and the target:
\begin{lstlisting}
letting t be 5
letting tgt be function ((1, 3) --> 1, (3, 1) --> 1,
  (2, 3) --> 1, (3, 4) --> 1, (3, 5) --> 1)
letting tlab be function (1 --> 1, 2 --> 1, 3 --> 1,
  4 --> 2, 5 --> 3)
\end{lstlisting}
Using the Conjure tool to compile Essence to a constraint programming model which is then solved by
Minion, we find there are exactly two solutions to the problem, as we would expect:
\begin{lstlisting}
(1 --> 1, 2 --> 3, 3 --> 4)
(1 --> 2, 2 --> 3, 3 --> 4)
\end{lstlisting}
But what if our application requires induced isomorphisms? Then we can easily add the constraint
\begin{lstlisting}
such that forAll (a, b) : (P, P) .
    (f(a), f(b)) in defined(tgt) -> (a, b) in defined(pat)
\end{lstlisting}
And Conjure will now find us a single solution,
\begin{lstlisting}
(1 --> 2, 2 --> 3, 3 --> 4)
\end{lstlisting}
As we will see in \cref{section:problems}, supporting other problem variants and constraints is
similarly straightforward, even if auxiliary variables are required. For example, if instead we want
to allow relabelling on vertex labels (which is typically not supported by dedicated solvers), we
could do the following:
\begin{lstlisting}
find r : function (total, injective) L --> L
such that forAll a : P .  r(plab(a)) = tlab(f(a))
\end{lstlisting}
and we would find two additional solutions,
\begin{lstlisting}
(1 --> 1, 2 --> 3, 3 --> 5)
(1 --> 2, 2 --> 3, 3 --> 5)
\end{lstlisting}
and if we removed the injective keyword for the relabelling, we would find a fifth mapping
\begin{lstlisting}
(1 --> 2, 2 --> 3, 3 --> 1)
\end{lstlisting}
We return to relabelling in \cref{section:typegraphs}.

@@

Unfortunately, whilst elegant and flexible, the performance of this approach leaves a lot to be
desired on basic subgraph isomorphism instances.  The computational experiments in this paper are
performed on a cluster of machines with dual Intel Xeon E5-2697A v4 processors and 512GBytes RAM
running Ubuntu 18.04.  We will be using the 14,621 unlabelled, undirected instances from
Solnon's benchmark suite\footnote{https://perso.liris.cnrs.fr/christine.solnon/SIP.html}. This
benchmark suite was originally designed for algorithm portfolios work
\cite{DBLP:conf/lion/KotthoffMS16}, and brings together several collections of application and
randomly-generated instances with varying difficulties and solution counts (including many
unsatisfiable instances). Some of the instances have up to 900 vertices and 14,420 edges in patterns
and up to 6,671 vertices and 209,000 edges in targets. These lead to rather large models, by
constraint programming standards: the largest generated table constraint has nearly half a million
entries. However, these sizes are realistic from an applications perspective, and it would be
desirable if solvers could handle even larger target graphs.

@@

In \cref{figure:solvers} we plot the cumulative number of instances solved over time for the
non-induced decision problem, comparing the high level approach to the Glasgow Subgraph Solver
\cite{DBLP:conf/gg/McCreeshP020} and PathLAD \cite{DBLP:conf/lion/KotthoffMS16} (the two strongest
CP-inspired approaches), and to VF2 \cite{DBLP:journals/pami/CordellaFSV04} and RI
\cite{DBLP:journals/bmcbi/BonniciGPSF13} (simpler algorithms which perform well on easy instances).
The high level approach has very slow startup times (which is to be expected as it involves
launching a Java virtual machine and reading in a very large table constraint), but much more
worryingly, only catches up with the worst other solver in number of instances solved as the timeout
approaches. Worse, as the scatter plot in \cref{figure:solvers} shows, there are almost no instances
where the high level approach does better than the Glasgow Subgraph Solver. (For induced problems,
the results are even less favourable.)

These first results motivate the remainder of this paper. We want to retain the convenience of the
high level modelling approach, and to be able to add arbitrary side constraints to suit different
applications, but we do not want to have to abandon the performance that dedicated solvers can get
on hard instances. In \cref{section:hybrid} we evaluate several ways of using a CP solver in
conjunction with the Glasgow Subgraph Solver, with a focus on low level implementation details. In
\cref{section:problems} we then return to high level modelling, and look at the convenience it
provides for retyping typed graphs, for temporal problems, and for optimisation problems.

\section{Hybrid Solving With High-Level Modelling}\label{section:hybrid}

Designing an effective hybrid solving system involved three major decisions: how the high-level
modelling language would identify suitable problems to hybridise, how the solvers would communicate,
and how often this communication would occur. The first two decisions were relatively
straightforward to make, but the third required using computational experiments to evaluate
different options. This section discusses all three of these decisions.

\subsection{High-Level Modelling}

We chose to use the Essence high-level constraint modelling language
\cite{DBLP:journals/constraints/FrischHJHM08} because of its support for convenient high-level types
like functions and relations. In a conventional modelling pipeline, problems are specified in
Essence and then are converted to concrete models that can be solved by a CP solver (in our
experiments Minion) via the Conjure and SavileRow tools. We augmented Conjure with a command line
option that instructs it to generate an extra file which describes how the variables representing
the graph are represented in the SavileRow input (known as Essence'). This is used by the graph
solver so it can map between its internal state and the variables in Essence'.  SavileRow converts
the graph model to Minion input, and this conversion contains information which allows Minion to map
its internal state back to the Essence' given to SavileRow. The graph solver and Minion then
communicate mappings between the nodes and edges in the graph using the identifiers in the Essence'
representation of the problem.

This design is based upon the notion that a CP solver and a subgraph solver can have enough of a
shared understanding of a problem to solve it co-operatively. Indeed, the Glasgow Subgraph Solver
\cite{DBLP:conf/gg/McCreeshP020} employs a CP approach to solve subgraph-finding problems, but using
special data structures and algorithms---for example, rather than representing the adjacency
constraint using an explicit table, it uses bitset adjacency matrices
\cite{DBLP:conf/cp/McCreeshP15}. The solver also exploits various graph invariants involving degrees
\cite{DBLP:journals/constraints/ZampelliDS10} and paths \cite{DBLP:conf/cp/AudemardLMGP14} to
further reduce the search space, and employs special search order heuristics
\cite{DBLP:conf/cpaior/ArchibaldDHMP019}. From this paper's perspective, the most important design
aspect is that internally, the solver has a CP style variable for each vertex in the pattern graph,
whose domains range over the vertices of the target graph. The solver performs a backtracking search
with restarts and nogood recording, attempting to assign each variable a value from its domain,
whilst respecting adjacency and injectivity constraints. At each recursive call of search, the
solver performs \emph{propagation} to eliminate infeasible values from domains. If any domain
becomes empty, the solver backtracks; otherwise, it selects a variable, and tries assigning it each
value from its domain in turn.

The high-level approach, then, gives us a way of setting up the subgraph solver and a CP solver such
that they both have an equivalent same set of variables and values for the graph part of the model,
and tells us how to form a correspondence between their internal representations. Importantly, this
allows the CP solver to have additional variables that the subgraph solver does not know about, and
we do not specifically require the CP solver to be aware of all of the graph constraints.
(Additionally, due to preprocessing, the CP solver may also sometimes have only a subset of values
for some graph variables visible to it.)

\subsection{When to Communicate?}

Having found a way to set up the two solvers, we must next ask when they should communicate. The
simplest approach would be to use the CP solver as a \emph{solution checker} for the subgraph
solver.  Whenever the subgraph solver finds a solution, it will pass it to the CP solver, which will
treat the solution as a set of equality constraints. The CP solver will then attempt to find a
satisfying assignment. If the CP solver does not have any additional variables, this is equivalent
to simply checking that the remaining constraints hold, but in general this will require search.
For a decision problem, the CP solver then communicates back to the subgraph solver either ``yes,
this is a valid solution'', or ``no, reject this solution and keep going''. If we are solving a
counting or enumeration problem, the CP solver must find \emph{all} solutions and communicate this
back to the subgraph solver.

For more power, but possibly also greater cost, we could additionally ask a CP solver at every stage
of search to test whether the subgraph solver is in an obviously infeasible state.  Whenever the
subgraph solver has finished performing propagation, it can communicate the \emph{trail} (that is,
its current sequence of guessed assignments) to the CP solver, which again treats these as
additional equality constraints. The CP solver then performs its own propagation (but not search),
and communicates back either a ``yes, keep going'' or a ``no, backtrack immediately''.  Finally,
after this testing, we could also ask the CP solver to communicate any deletions it infers back to
the subgraph solver. In other words, the subgraph solver would use the CP solver as an additional
propagator.

@@

\subsection{How to Communicate}

@@

We now present the results of some computational experiments. The experiments in this section are
designed to be hard, and to emphasise the difference between the approaches, rather
than to be realistic. We will continue to work with Solnon's non-induced subgraph isomorphism
benchmark instances, but will consider four variations. Firstly, we will consider the problem with
no side constraints. This is, in some sense, the worst case scenario, where we must pay the full
price of hybrid solving, but cannot get any benefit from it. Secondly, let us say that the number of
odd-indexed target vertices used must be greater than the number of even target vertices used:
\newcommand{\modsymbol}{\%}
\begin{lstlisting}
such that (sum i : P . f(i) (*\modsymbol{}*) 2) >
    (sum i : P . (f(i) + 1) (*\modsymbol{}*) 2)
\end{lstlisting}
Thirdly, let us instead say that fewer than three odd target vertices may be used:
\begin{lstlisting}
such that (sum i : P . f(i) (*\modsymbol{}*) 2) < 3
\end{lstlisting}
And fourthly, let us say that even pattern vertices must be mapped to even target vertices, and odd
pattern vertices to odd target vertices:
\begin{lstlisting}
such that forAll a : P . (a (*\modsymbol{}*) 2) = (f(a) (*\modsymbol{}*) 2)
\end{lstlisting}

We chose these problems because the second is likely not to be able to
perform inference until deep in a search tree (when most pattern vertices are mapped to specific
target vertices), the third is likely to be able to perform inference early in the search tree
(after a few assignments have been made, but not at the root node), and the fourth should propagate
only at the root node.

The results of these experiments are presented in \cref{figure:cumulative}. Let us first look at the
top left plot, where we do not actually have any side constraints. When calling the CP solver as a
solution checker, we ultimately achieve the same performance as the Glasgow Subgraph
Solver\footnote{Actually, because high level modelling can use different names for variables and
values, we get slight differences due to changes to tiebreaking in search order heuristics.},
although we can pay a substantial startup overhead. This should not be surprising: with no side
constraints, the subgraph solver runs as normal, and will perform just one call to the CP solver on
satisfiable instances. The propagating approach is over an order of magnitude slower: calling the CP
solver at every search node is clearly very expensive. (We also tried calling the CP solver to test
feasibility, without communicating deletions; this made no noticeable difference to performance, and
so is not pictured.) Finally, all approaches substantially outperform using a CP solver on its own
without help from the subgraph solver.

@@

As a point of comparison, we also implemented these three sets of side constraints natively inside
the Glasgow solver. For the ``more odd than even'' case, we did this through solution checking; for
``less than three odd'' we implemented checking during search; and for ``odd to odd, even to even''
we implemented initial domain filtering. Our implementation choices here are intended to reflect what a
reasonable programmer would do if the high level approach were not available (and we intentionally
selected side constraints that would not be too difficult to implement). In two of the three cases,
the hand-crafted code does somewhat outperform the hybrid solving, but in the ``more odd than even''
case, hybrid solving actually beats the hand-crafted dedicated solver implementation when using the
rollback approach, which we now describe.

@@

From what we have seen so far, it is obviously important to call the CP solver \emph{some} of the
time during search, but too expensive to call it \emph{all} of the time. We will therefore introduce
a new approach, which we call \emph{rollback}. This approach is inspired by backjumping
\cite{DBLP:journals/ci/Prosser93}, as well as by the conflict analysis methods used in SAT and SMT
solvers \cite{DBLP:conf/iccad/SilvaS96} and in lazy clause generating CP solvers
\cite{DBLP:conf/cp/OhrimenkoSC07,DBLP:conf/cpaior/Stuckey10}. The idea is as follows.  Firstly, we
call the CP solver with full propagation at the root node, in case we are dealing with a
particularly rich labelling scheme.  Secondly, we use the CP solver for solution checking, since
this is required for correctness. Now, suppose the CP solver rejects a candidate solution: this will
cause the subgraph solver to backtrack. At this point, we call the CP solver again, with full
propagation. Either the CP solver indicates feasibility, in which case we proceed with search
(potentially with a reduced set of domains), or the CP solver indicates failure, in which case we
backtrack again, and do another attempt at full propagation, and so on until feasibility is reached.

@@

We now look at three classes of real-world subgraph-finding problems that, until now, have been
solved using dedicated approaches. We show how easy it is to model these problems in Essence,
demonstrating the usefulness of the high-level modelling approach for prototyping and development.

@@

The basic notion of a graph conveys only adjacency information, and a subgraph isomorphism simply
finds a certain structural pattern. In practice, this is often augmented with additional
information---for example, we have seen how labels can be associated with vertices and edges, which
can be used in chemical applications to represent different kinds of atom or bond. In this case,
subgraph isomorphisms are also expected to preserve labels, so carbon atoms can only be mapped to
carbon atoms, and double bonds must be mapped to double bonds.  A richer labelling abstraction comes
in the form of \emph{typed graphs}, where the labels themselves also carry a graph structure
\cite{Ehrig+06}; we show an example in \cref{figure:littlegraphs}. In practice this labelling
structure is specified by providing two graphs, together with a morphism from the main graph to its
type graph.

For typed graphs, morphisms between the graphs are typically defined between graphs typed over the
same type graph, but there are situations where we are interested in mapping between graphs typed
over different type graphs.  One such scenario from a software engineering context is described by
Dur{\'a}n et al.~\cite{Duran+12,Duran+17}, where graph transformation systems are composed by
defining morphisms between the rules constituting the respective transformation systems.  In this
case, the source and target transformation systems will normally have different type graphs, a
morphism must also be established between the two type graphs.  Mappings between the various graphs
making up the rules then need to preserve structure and typing subject to the morphism between type
graphs. This approach to specification composition is implemented by the gts-morpher
tool.\footnote{\url{https://github.com/gts-morpher/gts_morpher}} A key objective is to minimise the
amount of specification that needs to be written.  For example, the tool allows morphisms between
graph transformation systems to be only partially specified and then automatically completes the
full morphism, if it can do so unambiguously---this requires solving a sub-isomorphism problem.

We may describe typed graph subisomorphism problems in Essence as follows. As before, we are given a
pattern graph and a target graph, both of which carry labels; we will draw the vertex labels from
different sets, to emphasise
the relabelling.
\begin{lstlisting}
given pl, tl, e : int
letting PL be domain int(1..pl)
letting TL be domain int(1..tl)
letting E be domain int(1..e)
\end{lstlisting}
We are also given labelled graphs,
\begin{lstlisting}
given p, t : int
letting P be domain int(1..p)
letting T be domain int(1..t)

given pat : function (P, P) --> E
given tgt : function (T, T) --> E
given plab : function (total) P --> PL
given tlab : function (total) T --> TL
\end{lstlisting}
but now the labels also carry a graph structure,
\begin{lstlisting}
given pattype : function (PL, PL) --> E
given tgttype : function (TL, TL) --> E
\end{lstlisting}
We are looking for an injective mapping from the pattern graph to the target graph,
\begin{lstlisting}
find f : function
    (total, injective) P --> T
\end{lstlisting}
as well as an injective mapping between the label graphs,
\begin{lstlisting}
find r : function
    (total, injective) PL --> TL
\end{lstlisting}
in such a way that graph structure and labels are preserved,
\begin{lstlisting}
such that forAll ((a, b), lbl) in pat .
    ((f(a), f(b)), lbl) in toSet(tgt)
such that forAll a : P .
    r(plab(a)) = tlab(f(a))
\end{lstlisting}
and also requiring that the structure on the labels is preserved,
\begin{lstlisting}
such that forAll (a,b) in defined(pattype) .
    pattype((a,b)) = tgttype((r(a),r(b)))
\end{lstlisting}
